---
title: "example_plink"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example_plink}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we'll use GCAE on the PLINK tutorial's example data, 
to train and use a GCAE neural network.

Here are the steps:

 * Setup GCAE
 * See the genetic data
 * See the GCAE model
 * Use GCAE for training

# Setup GCAE

First we determine if GCAE is installed 
and if the PLINK tutorial data is installed.

To determine if GCAE is installed, we'll first load `gcaer`:

```{r setup}
library(gcaer)
```

Now, `gcaer` detects if GCAE is installed:

```{r is_gcae_installed}
if (!is_gcae_installed()) {
  message("GCAE is not installed")
  message("Tip: use 'gcaer::install_gcae()'")
}
```

Here we use `plinkr` to detect if the PLINK tutorial data is installed:

```{r is_plink_tutorial_data_installed}
if (!plinkr::is_plink_tutorial_data_installed()) {
  message("PLINK tutorial data is not installed")
  message("Tip: use 'plinkr::install_plink_tutorial_data'")
}
```

If both requirements are met, we are good to go:

```{r good_to_go}
good_to_go <- is_gcae_installed() && plinkr::is_plink_tutorial_data_installed()
```

Without these requirements met, this vignette will be rather dull :-)

## See the genetic data

These are the PLINK tutorial files:

```{r all_plink_tutorial_data_filenames}
if (good_to_go) {
  plink_tutorial_data_filenames <- plinkr::get_plink_tutorial_data_filenames()
  plink_tutorial_data_filenames
}
```


We will be using `hapmap1` files, 
that are in the non-binary/plain-text PLINK file format:

```{r hapmap1_tutorial_data_filenames}
if (good_to_go) {
  plain_text_plink_filenames <- stringr::str_subset(
    string = plink_tutorial_data_filenames,
    pattern = "hapmap1"
  )
  plain_text_plink_filenames
}
```

As GCAE needs these files to be in binary PLINK format, we convert the
text files to binary files and store these in a temporary folder:

```{r make_bed}
if (good_to_go) {
  temp_folder <- plinkr::get_plinkr_tempfilename()
  binary_plink_filenames <- plinkr::make_bed(
    base_input_filename = tools::file_path_sans_ext(
      plain_text_plink_filenames[1]
    ),
    base_output_filename = file.path(temp_folder, "hapmap1_binary")
  )
  binary_plink_filenames
}
```

The three PLINK binary files are the `.fam` and `.bim`, `.bed` files.

This is the `.fam` file:

```{r read_fam_file}
n_individuals <- "[unknown]"
if (good_to_go) {
  fam_filename <- stringr::str_subset(
    binary_plink_filenames,
    pattern = "\\.fam$"
  )
  fam_table <- plinkr::read_plink_fam_file(fam_filename)
  n_individuals <- nrow(fam_table)
  knitr::kable(utils::head(fam_table))
}
```

This table shows us the `r n_individuals` individuals.

This is the `.bim` file:

```{r read_bim_file}
n_snps <- "[unknown]"
if (good_to_go) {
  bim_filename <- stringr::str_subset(
    binary_plink_filenames,
    pattern = "\\.bim$"
  )
  bim_table <- plinkr::read_plink_bim_file(bim_filename)
  n_snps <- nrow(bim_table)
  knitr::kable(utils::head(bim_table))
}
```

The `.bim` file shows the information of the `r n_snps` SNPs.

The `.bed` file connects the individuals and the SNPs:

```{r read_bed_file}
if (good_to_go) {
  bed_filename <- stringr::str_subset(
    binary_plink_filenames,
    pattern = "\\.bed$"
  )
  bed_table <- plinkr::read_plink_bed_file(
    bed_filename,
    names_loci = bim_table$id,
    names_ind = fam_table$fam
  )
  testthat::expect_equal(ncol(bed_table), n_individuals)
  testthat::expect_equal(nrow(bed_table), n_snps)
  knitr::kable(utils::head(bed_table[, 1:10]))
}
```

## See the GCAE model


## Use GCAE for training
