---
title: "example_tiny"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{example_tiny}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this vignette, we'll use GCAE and its `example_tiny` example data, to train and use a GCAE neural network.

Here are the steps:

 * Setup GCAE
 * See the genetic data
 * See the GCAE model
 * Use GCAE for training

# Setup GCAE

First we determine if GCAE is installed.

To determine if GCAE is installed, we'll first load `gcaer`:

```{r setup}
library(gcaer)
```

Now `gcaer` can detect if GCAE is installed:

```{r is_gcae_installed}
good_to_go <- is_gcae_installed()
if (!good_to_go) {
  message("GCAE is not installed")
  message("Tip: use 'gcaer::install_gcae()'")
}
```

## See the genetic data

These are the `example_tiny` files supplied with GCAE:

```{r data_dir}
if (good_to_go) {
  data_dir <- file.path(get_gcae_subfolder(), "example_tiny/")
  list.files(data_dir)
}
```
Of these files, we will use `HumanOrigins249_tiny` as the base file name:

```{r base_filename}
if (good_to_go) {
  base_filename <- "HumanOrigins249_tiny"
  list.files(data_dir, pattern = base_filename)
}
```

We will be using the PLINK binary files, which are the `.bed`, `.bim` 
and `.fam` files.

This is the `.fam` file:

```{r read_fam_file}
n_individuals <- "[unknown]"
if (good_to_go) {
  fam_filename <- list.files(data_dir, full.names = TRUE, pattern = "\\.fam$")
  fam_table <- plinkr::read_plink_fam_file(fam_filename)
  n_individuals <- nrow(fam_table)
  knitr::kable(head(fam_table))
}
```

This table shows us the `r n_individuals` individuals.

This is the `.bim` file:

```{r read_bim_file}
if (good_to_go) {
  bim_filename <- list.files(data_dir, full.names = TRUE, pattern = "\\.bim$")
  bim_table <- plinkr::read_plink_bim_file(bim_filename)
  knitr::kable(head(bim_table))
}
```

This is the `.bed` file:

```{r read_bed_file}
if (good_to_go) {
  bed_filename <- list.files(
    data_dir, 
    full.names = TRUE, 
    pattern = "\\.bed$"
  )
  bed_table <- plinkr::read_plink_bed_file_from_files(
    bed_filename = bed_filename,
    bim_filename = bim_filename,
    fam_filename = fam_filename
  )
  knitr::kable(head(bed_table[, 1:10]))
}
```

## See the GCAE model

### The model ID

```{r}
model_id <- "M1"
if (good_to_go) {
  model_filename <- file.path(get_gcae_subfolder(), "models", paste0(model_id, ".json"))
  testthat::expect_true(file.exists(model_filename))
  readr::read_lines(model_filename)
}
```

### train_opts_id

```{r}
train_opts_id <- "ex3"
if (good_to_go) {
  train_opts_filename <- file.path(get_gcae_subfolder(), "train_opts", paste0(train_opts_id, ".json"))
  testthat::expect_true(file.exists(train_opts_filename))
  readr::read_lines(train_opts_filename)
}
```

###

### train_opts_id

```{r}
data_opts_id <- "b_0_4"
if (good_to_go) {
  data_opts_filename <- file.path(get_gcae_subfolder(), "data_opts", paste0(data_opts_id, ".json"))
  testthat::expect_true(file.exists(data_opts_filename))
  readr::read_lines(data_opts_filename)
}
```
## Use GCAE for training

```{r}
if (good_to_go) {
  gcae_train(
    datadir = data_dir,
    data = base_filename,
    model_id = model_id,
    epochs = 1,
    save_interval = 1,
    train_opts_id = train_opts_id,
    data_opts_id = data_opts_id,
    gcae_options = create_gcae_options(),
    verbose = FALSE
  )
}
```



